The ChEMBL database as Linked Open Data

### 2012-05-28 article.tex
More tuning: added two references, unnumbered sections

The current scientific data deluge, in which datasets are growing faster than scientists can analyse and 
curate them, provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets, which explaining biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independent of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{FIXME}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,OpenPHACTS}.

@@

\subsection{ChEMBL-RDF}

\subsubsection{Data Structure}

@@

\subsubsection{Data Statistics and Validation}

Each release of the ChEMBL database is accompanied by set of release counts, which help 
provide a concise content overview of the release in question. The ChEMBL 13 release counts
are as follows:

\begin{itemize}
  \item 8,845 targets
  \item 1,143,682 distinct molecules
  \item 6,933,068 activities
  \item 617,681 assays
  \item 44,682 documents
\end{itemize}

@@

\begin{tiny}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Target .
}
\end{verbatim}
\end{tiny}

@@

\begin{tiny}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Molecule .
}
\end{verbatim}
\end{tiny}

@@

\begin{tiny}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Assay .
}
\end{verbatim}
\end{tiny}

@@

\begin{tiny}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Activity .
}
\end{verbatim}
\end{tiny}

@@

\subsection{Linked Open Data}

@@

\subsubsection{Linking out to Bio2RDF}

@@

\begin{tiny}
\begin{verbatim}
target:t101191
 chembl:hasTaxonomy <http://bio2rdf.org/taxonomy:9606> ;
 owl:sameAs <http://bio2rdf.org/ec:2.7.11.1> ;
 owl:sameAs <http://bio2rdf.org/uniprot:Q8IYT8> .
\end{verbatim}
\end{tiny}

@@

\begin{tiny}
\begin{verbatim}
res:r23 skos:exactMatch <http://bio2rdf.org/pubmed:15149661> .
\end{verbatim}
\end{tiny}

\subsubsection{Linking out to ChemSpider}

@@

\begin{tiny}
\begin{verbatim}
<http://data.kasabi.com/dataset/chembl-rdf/chemblid/ChEMBL324846>
 skos:exactMatch <http://rdf.chemspider.com/370> .
\end{verbatim}
\end{tiny}

\subsubsection{Linking out to OpenMolecules RDF}

@@

\begin{tiny}
\begin{verbatim}
mol:m62687 owl:equivalentClass
 <http://rdf.openmolecules.net/?InChI=1S/CH4/h1H4> .
\end{verbatim}
\end{tiny}

@@

\subsubsection{Linking out to CrossRef}

@@

\begin{tiny}
\begin{verbatim}
res:r2032 owl:sameAs <http://dx.doi.org/10.1016/0960-894X(96)00111-4> .
\end{verbatim}
\end{tiny}

@@

\subsection{Bio2RDF}

@@

\subsection{CitedIn}

@@

\subsection{Bioclipse Decision Support}

@@

\subsection{Chem2Bio2RDF}

@@

\subsection{Compound Selectivity}

@@

While we show here that the RDF version of the ChEMBL data is very useful, the current RDF triples are by no means static: ChEMBL-RDF
will keep evolve, following new developments in the Linked Open Data world. For example, it is likely that over time we will link out
to more Linked Data resources. More importantly, we wish to adopt more common ontologies, of which the BioAssay Ontology is planned
to be the next~\cite{Visser2011}. These steps will make the ChEMBL-RDF triples even more interoperable.

>>>>

The current scientific data deluge, in which datasets are growing faster than scientists can analyse and 
curate them, provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets, which explaining biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independent of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{Staab2007}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,OpenPHACTS}.

@@

\subsection*{ChEMBL-RDF}

\subsubsection*{Data Structure}

@@

\subsubsection*{Data Statistics and Validation}

Each release of the ChEMBL database is accompanied by set of release counts, which help 
provide a concise content overview of the release in question. The ChEMBL 13 release counts
are as follows:
\begin{itemize}
  \item 8,845 targets
  \item 1,143,682 distinct molecules
  \item 6,933,068 activities
  \item 617,681 assays
  \item 44,682 documents
\end{itemize}

@@

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Target .
}
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Molecule .
}
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Assay .
}
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Activity .
}
\end{verbatim}
\end{small}

@@

\subsection*{Linked Open Data}

@@

\subsubsection*{Linking out to Bio2RDF}

@@

\begin{small}
\begin{verbatim}
target:t101191
 chembl:hasTaxonomy <http://bio2rdf.org/taxonomy:9606> ;
 owl:sameAs <http://bio2rdf.org/ec:2.7.11.1> ;
 owl:sameAs <http://bio2rdf.org/uniprot:Q8IYT8> .
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
res:r23 skos:exactMatch <http://bio2rdf.org/pubmed:15149661> .
\end{verbatim}
\end{small}

\subsubsection*{Linking out to ChemSpider}

@@

\begin{small}
\begin{verbatim}
<http://data.kasabi.com/dataset/chembl-rdf/chemblid/ChEMBL324846>
 skos:exactMatch <http://rdf.chemspider.com/370> .
\end{verbatim}
\end{small}

\subsubsection*{Linking out to OpenMolecules RDF}

@@

\begin{small}
\begin{verbatim}
mol:m62687 owl:equivalentClass
 <http://rdf.openmolecules.net/?InChI=1S/CH4/h1H4> .
\end{verbatim}
\end{small}

@@

\subsubsection*{Linking out to CrossRef}

@@

\begin{small}
\begin{verbatim}
res:r2032 owl:sameAs <http://dx.doi.org/10.1016/0960-894X(96)00111-4> .
\end{verbatim}
\end{small}

@@

\subsection*{Bio2RDF}

@@

\subsection*{CitedIn}

@@

\subsection*{Bioclipse Decision Support}

@@

\subsection*{Chem2Bio2RDF}

@@

\subsection*{Compound Selectivity}

@@

While we show here that the RDF version of the ChEMBL data is very useful, the current RDF triples are by no means static: ChEMBL-RDF
will keep evolve, following new developments in the Linked Open Data world. For example, it is likely that over time we will link out
to more Linked Data resources. More importantly, we wish to adopt more common ontologies, of which the BioAssay Ontology is planned
to be the next~\cite{Visser2011}. These steps will make the ChEMBL-RDF triples even more interoperable. Part of this future work
is already being performed within the context of the Open PHACTS project~\cite{Williams2012}.
